---
title: "kallisto h5 files to tximport to DESeq2 Walkthrough"
author: "Aaron Mitchell-Dick"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Users/AMDprograms/Desktop/Untrimmed/')
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(error = FALSE)
```

<br/>
<br/>

# RNA-seq analysis using kallisto, tximport, and DeSeq2

This walkthrough was built by combining the information from several vignettes in order to streamline workflow from kallisto output files.  
To see the beginning of the workflow from raw sequencing reads (fastq files) to finished kallisto output directory, see [Raw Sequencing Reads -> filtering(FastQC, TrimGalore) -> kallisto](https://aaronmitchd.github.io/RNA_Seq/raw_seq_reads_to_kallisto.html).  
Here, we start with `.h5` files already ordered into a directory, and we have already created a `.txt` files that describes the experimental paradigm, including samples and conditions.  
If you are executing this for the very first time, see this file [Install Packages](https://aaronmitchd.github.io/RNA_Seq/Install-Packages.html) for information on how to download all the packages you will need, which you'll need to do first before you can implement the function `library()`.

<br/>

#### Load your libraries:

```{r libraries,message=FALSE}
library("tximport")
library("readr")
library("tximportData")
library("rhdf5")
library("GenomicFeatures")
library("DESeq2")
library("pacman")
```

<br/>

#### Set up your working directory

Set the file path to the folder that contains: 
1) A directory of folders wtih each of your abundance.h5 files from the kallisto operation, and contains 
2) The .txt file describing samples and conditions

```{r directory, eval=FALSE}
setwd("C:/Users/AMDprograms/Desktop/Untrimmed/")
```

Turn the folder contents into a dataframe called dir:

```{r dir df,message=FALSE}
dir <- dir(file.path("./"))
head(dir)
```

<br/>

#### Make a dataframe that describes your experimental conditions
Turn the list of sample names into a dataframe:

```{r samples df}
samples <- read.table(file.path("./samples.txt"), header=TRUE)
samples
```

Make a dataframe with rows listed by sample name:

```{r samples rows}
rownames(samples) <- samples$sample
samples
```

Make a dataframe that lists the full path to all the kallisto .h5 files:

```{r .h5 file path}
files <- file.path("./kallisto", samples$sample, "abundance.h5")
files
```

Add column headers to the files dataframe ordered `sample[+n]`. All this does is provide columns for each sample's reads:

```{r smaples to rows}
names(files) <- paste0("sample", 1:6)
files
```

<br/>

#### Making the TxDb database to relate transcript name to gene name 

Getting the right mapping for turning your kallisto list of transcripts into a list useable for tximport is important, and somewhat tricky. A few different vignettes suggested codes that did not work for me.  Luckily, the authors of kallisto maintain [a database of tarball downloads](https://github.com/pachterlab/kallisto-transcriptome-indices/releases) that not only contain high-quality pre-made kallisto indexes (.idx files) for common organisms, but also contain a `.gtf` file which is exactly what you can use here to make the TxDb object. So, in short, I used the kallisto tarball which includes the organism's gtf file as input into TxDb, making it into a TxDb dataframe.
Simply navigate to the directory and file where you extracted the tarball (via Unix/Linux, on Windows Linux using WSL/Ubuntu, or on Mac) and you're good to go.

```{r tarball gtf,message=FALSE}
txdb <- makeTxDbFromGFF("C:/Users/AMDprograms/Downloads/mus_musculus/mus_musculus/Mus_musculus.GRCm38.96.gtf")
```

You can check what this data is like by calling these:

```{r check}
columns(txdb)
keytypes(txdb)
```

Make a key of the transcript names, make a dataframe of this key.

```{r keymaker}
k <- keys(txdb, keytype = "TXNAME")
```

Make a dataframe using that key to map transcript names to gene names.  This will allow tximport to run.

```{r tx map}
tx_map <- select(txdb, keys = k, columns="GENEID", keytype="TXNAME")
head(tx_map)
```

<br/>

#### Run tximport

Run tximport, and outputs to a dataframe txi.kallisto:

```{r tximport}
#convert dataframe to new name
tx2gene <- tx_map
# then
txi.kallisto <- tximport(files, type = "kallisto", tx2gene=tx2gene, ignoreTxVersion = TRUE)
head(txi.kallisto$counts)
```

<br/>

#### Make a DESeq2 object

Turn the tximport dataset into a DeSeq2 object, called a DeSeqDataset:

```{r txi to dds,message=FALSE}
ddsTxi <- DESeqDataSetFromTximport(txi.kallisto,
                                   colData = samples,
                                   design = ~ condition)
ddsTxi
```

<br/>

#### Prefilter the reads to remove low read counts

This prefiltering step is suggested in the DESeq2 vignette.
It removes any rows that have less than a total of 10 reads across all samples:

```{r prefilter}
keep <- rowSums(counts(ddsTxi)) >= 10
dds <- ddsTxi[keep,]
dds
# the above prefiltering function got rid of many transcripts (36047-x=20185)
```

<br/>

#### Make sure DESeq2 compares the correct experimental conditions.

This next step will just make sure comparisons are made to the correct control. name the ref to whatever your control condition is named (here it's DMSO):

```{r set comparison}
dds$condition <- relevel(dds$condition, ref = "DMSO")
```

<br/>

#### Run DESeq2

Run DeSeq2, which returns a results file `res` with log2 fold changes, p values and adjusted p values.

```{r DeSeq2, message=FALSE}
dds <- DESeq(dds)
res <- results(dds)
head(res)
```

this checks and tells you the comparison that was made (e.g. experimental vs control):

```{r rightcomparison}
resultsNames(dds)
```

<br/>

#### (Optional) Shrink effect size for making graphs
This is going to shrink effect size, using apelgm estimator (see comment in the block for why this is used):

```{r shrink, eval=FALSE}
resFC <- lfcShrink(dds, coef="condition_STLC_vs_DMSO", type="apeglm")
head(resLFC)
# Shrinkage is not needed for text-based analysis, So the estimator
# output is written to a different variable.  This would be useful for graphing.
```

<br/>

#### Examine the output of the differential expression analysis

Order your data list in any way you'd like, here it's ordered by pvalue:

```{r order}
resOrdered <- res[order(res$pvalue),]
summary(res)
```

You can filter your summary by any parameters you'd like, here its padj <0.1:

```{r cutoff}
sum(res$padj < 0.1, na.rm=TRUE)
```

Write your results to a file:

```{r write, eval=FALSE}
write.csv(as.data.frame(resOrdered), 
          file="0h_STLC_vs_DMSO.csv")
#you can write this file to output but it is not as detailed as the file 
#that is output below, which contains gene descriptions.
```

Check if all mapping of the data is good:

```{r mapping correct}
#check if all mapping of the data is good
any(duplicated(rownames(res)))
any(duplicated(colnames(res)))
```

<br/>

#### Add common gene names and short descriptions to your DESeq2 output

load these libraries to prepare to add gene annotations to the dataset:

```{r load,message=FALSE}
library("AnnotationDbi")
library("org.Mm.eg.db")
```

Append gene names (symbols) to the results file:

```{r naming}
ens.str2 <- rownames(resOrdered)
head(ens.str2)
resOrdered$symbol <- mapIds(org.Mm.eg.db,
                     keys=ens.str2,
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
head(resOrdered)
```


this turns the resOrdered object into a dataframe in prep for writing to csv file.

```{r convert df}
resOrdered_df <- as.data.frame(resOrdered)
```

<br/>

#### Write your analysis to a .csv file

This now executes a block which does:
1) converts our DESeq2 results to a dataframe
2) makes a merged table that sums together gene names and descriptions

```{r merge and write final}
# this will make a dataframe out of merged gene names and shrot descriptions.
gene_desc <- merge(org.Mm.egSYMBOL, org.Mm.egGENENAME)
# we then create a new dataframe merging the gene descriptions to the modified DeSeq2 output.
res_desc <- merge(resOrdered_df, gene_desc)
head(res_desc)
```

this writes a file that contains gene names, DiffEx data, stats, and gene descriptions all together to the current working directory.

```{r final write, eval=FALSE}
write.csv(res_desc, 
          file="gene_desc_0h_STLC_vs_DMSO.csv")
```

Your data is now readable in R, command line, python, excel, etc.!
